---
title: "fmtr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fmtr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **fmtr** package helps format data.  The package aims to simulate 
the basic functionality of SAS® formats, but with R.  

Formatting may be assigned
to data frame columns using the **format**, **width**, and **justify** 
attributes.  Formatting is then applied by calling the `fdata` function. 
`fdata` returns a new data frame with the specified formatting applied. This 
method of formatting provides much greater control over data formatting
than the base R `format` function.

Here is a example:

```{r eval=FALSE, echo=TRUE}
# Construct data frame from state vectors
df <- data.frame(state = state.abb, area = state.area)[1:10, ]

# Calculate percentages
df$pct <- df$area / sum(state.area) * 100

# Before formatting 
df

#    state   area         pct
# 1     AL  51609  1.42629378
# 2     AK 589757 16.29883824
# 3     AZ 113909  3.14804973
# 4     AR  53104  1.46761040
# 5     CA 158693  4.38572418
# 6     CO 104247  2.88102556
# 7     CT   5009  0.13843139
# 8     DE   2057  0.05684835
# 9     FL  58560  1.61839532
# 10    GA  58876  1.62712846

# Create state name lookup list
name_lookup <- state.name
names(name_lookup) <- state.abb

# Assign formats
formats(df) <- list(state = name_lookup,                         
                    area  = function(x) format(x, big.mark = ","), 
                    pct   = "%.1f%%") 

# Apply formats
fdata(df)

#          state    area   pct
# 1      Alabama  51,609  1.4%
# 2       Alaska 589,757 16.3%
# 3      Arizona 113,909  3.1%
# 4     Arkansas  53,104  1.5%
# 5   California 158,693  4.4%
# 6     Colorado 104,247  2.9%
# 7  Connecticut   5,009  0.1%
# 8     Delaware   2,057  0.1%
# 9      Florida  58,560  1.6%
# 10     Georgia  58,876  1.6%

```
## Key Features

**fmtr** contains the following key features:

* The `fdata` function to apply formatting to any data frame or tibble.
* The `fapply` function to apply formatting to any vector.
* The `formats` and `fattr` functions to easily assign formatting attributes.
* The `value` and `condition` functions to create a new user-defined format.
* The `fcat` function to create a format catalog (NEW in v1.2).
* The `flist` function to create a formatting list.

