---
title: "fmtr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fmtr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **fmtr** package helps format data.  The package aims to simulate 
the basic functionality of SAS® formats, but with R.  The package contains
several functions that make formatting simpler and more powerful.

### Key Functions

**fmtr** contains the following key functions:

* The `fdata()` function to apply formatting to any data frame or tibble.
* The `fapply()` function to apply formatting to any vector.
* The `formats()` and `fattr()` functions to easily assign formatting attributes.
* The `value()` and `condition()` functions to create a user-defined format.
* The `fcat()` function to create a format catalog.
* The `flist()` function to create a formatting list.

### How to Use
The **fmtr** package builds heavily on existing R formatting capabilities.
For most R programmers, these functions are well-known, and widely used.
The examples below make use of standard R formatting codes, such as 
those associated with the `strptime()` and `sprintf()` functions.  The
standard R formatting codes are a flexible and compact way of defining 
a format.  If you are unfamiliar with R formatting codes, please see
this summary on the `FormattingStrings()` page.


#### The `fapply()` Function
The simplest way to introduce the **fmtr** package is to examine the use of
the `fapply()` function.  

```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Create sample data vector
v1 <- c(1.483, 5.29837, 7.9472, 8.684021)

# Apply format
fapply(v1, "%1.1f")
# [1] "1.5" "5.3" "7.9" "8.7"
```
As you can see from the above example, the `fapply()` function typically
takes two parameters: a vector and a format.  In this way, the `fapply()` acts
very much like a SAS® `put` function. 

Note that the format parameter can also be assigned as an attribute on 
the vector.  The `fapply()` function will then pick up the format attribute,
and apply it to the input vector.  The result is the same:

```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Create sample data vector
v1 <- c(1.483, 5.29837, 7.9472, 8.684021)

# Assign format attribute
attr(v1, "format") <- "%1.1f"

# Apply format
fapply(v1)
# [1] "1.5" "5.3" "7.9" "8.7"
```

Besides the format attribute, the `fapply()` function will also recognize
attributes for `width` and `justify`.  These parameters allow you
to control the width and alignment of the data in the vector.  If the 
width parameter is larger than the width of the data, the value will be 
padded with spaces.  Here is an example:

```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Create sample data vector
v1 <- c(1.483, 5.29837, 7.9472, 8.684021)

# Assign formatting attributes
attr(v1, "format") <- "%1.1f"
attr(v1, "width") <- 5
attr(v1, "justify") <- "right"

# Apply formatting attributes
fapply(v1)
# [1] "  1.5" "  5.3" "  7.9" "  8.7"
```

To help simplify assignment of these attributes, the **fmtr** package
includes the `fattr()` function, which allows you to set all the above
attributes in one function call.  Here is an example using the `fattr()`
function, that ends with the same result as the example above.
```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Create sample data vector
v1 <- c(1.483, 5.29837, 7.9472, 8.684021)

# Assign formatting attributes
v1 <- fattr(v1, format = "%1.1f", width = 5, justify = "right")

# Apply formatting attributes
fapply(v1)
# [1] "  1.5" "  5.3" "  7.9" "  8.7"
```

Note that `fapply()` can accept several different types of formats.
The examples above focus on a simple numeric format.  But `fapply()` also 
accepts date formats, a lookup list, a user-defined format, 
a vectorized function, and a formatting list.  

Here is an example showing the use of a lookup list:
```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Create sample data vector
v1 <- c("A", "B", "A", "C", "B")

# Create lookup vector
v2 <- c(A = "Group A", B = "Group B", C = "Group C")

fapply(v1, v2)
# [1] "Group A" "Group B" "Group A" "Group C" "Group B"
```

#### A User-Defined Format

The weakness with using a named vector as a lookup list, 
as in the above example, is that there is no way to include any sort of logic
in the lookup.  For instance, if your data has NA values, you may want to handle
those differently from the valid input values. Or you may want to define 
a default value if the input data does not match any of the lookup keys.

For these reasons, the **fmtr** package
provides a *user-defined format*.  This concept was taken directly from 
SAS® software. The functions that create a user-defined format are 
`value()` and `condition()`. 

A condition accepts an expression and a label.
The expression determines which label is assigned.  For the expression, 
you can use logical operators like "&" and "|", and relational operators
like ">" and "<".  The data value is identified with a variable "x". Here is
an example:
```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Create sample data vector
v1 <- c("A", "B", "E", "A", NA, "C", "D")

u1 <- value(condition(x == "A", "Group A"),
            condition(x == "B", "Group B"),
            condition(x == "C" | x == "D", "Group C/D"), 
            condition(TRUE, "Other"))
            
fapply(v1, u1)  
# [1] "Group A" "Group B" "Other" "Group A" "Other" "Group C/D" "Group C/D"
```
Notice that the user-defined format gives you 
much more capabilities than a simple lookup vector.  It allows you to 
perform categorization, and assign a default.  Additionally, the NA missing value
does not crash the function.  The NA simply falls into the default category.
If there is no default category, any values which do not correspond to a 
category will fall through the format unaltered.

#### The `fdata()` Function

The `fdata()` function works very much the same way as `fapply()`, but
with data frames and tibbles instead of vectors.  In fact, under the hood, 
`fdata()` is simply calling `fapply()` for each column in the data frame.

Like the `fapply()` function, formatting may be assigned
to data frame columns using the **format**, **width**, and **justify** 
attributes.  Formatting is then applied by calling the `fdata()` function, and
passing the data frame as the first parameter.
`fdata()` will then return a new data frame with the specified formatting applied. 
This method of formatting provides much greater control 
than the base R `format()` function.

```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Construct data frame from state vectors
df <- data.frame(state = state.abb, area = state.area)[1:10, ]

# Calculate percentages
df$pct <- df$area / sum(state.area) * 100

# Before formatting 
df
#    state   area         pct
# 1     AL  51609  1.42629378
# 2     AK 589757 16.29883824
# 3     AZ 113909  3.14804973
# 4     AR  53104  1.46761040
# 5     CA 158693  4.38572418
# 6     CO 104247  2.88102556
# 7     CT   5009  0.13843139
# 8     DE   2057  0.05684835
# 9     FL  58560  1.61839532
# 10    GA  58876  1.62712846

# Create state name lookup list
name_lookup <- state.name
names(name_lookup) <- state.abb

# Assign formats
formats(df) <- list(state = name_lookup,                         
                    area  = function(x) format(x, big.mark = ","), 
                    pct   = "%.1f%%") 

# Apply formats
fdata(df)
#          state    area   pct
# 1      Alabama  51,609  1.4%
# 2       Alaska 589,757 16.3%
# 3      Arizona 113,909  3.1%
# 4     Arkansas  53,104  1.5%
# 5   California 158,693  4.4%
# 6     Colorado 104,247  2.9%
# 7  Connecticut   5,009  0.1%
# 8     Delaware   2,057  0.1%
# 9      Florida  58,560  1.6%
# 10     Georgia  58,876  1.6%

```
In the above example, observe that the `formats()` function assigns
the format attribute for multiple columns.  This assignment is accomplished
by sending a named list into the `formats()` function, where the names 
in the list correspond to the column names of the data frame.  Also note the use
of a lookup style format for the state names, and an anonymous vectorized 
format function for the state area.

#### The `fcat()` Function

One of the benefits of the above method of formatting is that the 
data frame attributes can be stored with the data frame, and reapplied
in the future.  But what if you want to apply the same set of formats to a 
different data frame?

That is where you need a *format catalog*.  

The format catalog is a collection of formats that can be saved and reused.
A format catalog is created with an `fcat()` function.  To create a format
catalog, you call the `fcat()` function, passing a set of name/format pairs.
In this case, the name of the format is a generic format name.  It does not
have to correspond to a column name.  You may name the formats anything 
you want.  The formats can be accessed in the catalog using dollar sign ("$")
list notation.
```{r eval=FALSE, echo=TRUE}
library(fmtr)

# Construct data frame from state vectors
df <- data.frame(state = state.abb, area = state.area)[1:10, ]

# Calculate percentages
df$pct <- df$area / sum(state.area) * 100

# Before formatting 
df
#    state   area         pct
# 1     AL  51609  1.42629378
# 2     AK 589757 16.29883824
# 3     AZ 113909  3.14804973
# 4     AR  53104  1.46761040
# 5     CA 158693  4.38572418
# 6     CO 104247  2.88102556
# 7     CT   5009  0.13843139
# 8     DE   2057  0.05684835
# 9     FL  58560  1.61839532
# 10    GA  58876  1.62712846

# Create state name lookup list
name_lookup <- state.name
names(name_lookup) <- state.abb

# Assign formats to format catalog
cat1 <- fcat(state = name_lookup,                         
             area  = function(x) format(x, big.mark = ","), 
             pct   = "%.1f%%") 
             
# Apply a format from the catalog using fapply
fapply(df$pct, cat1$pct)
# [1] "1.4%"  "16.3%" "3.1%"  "1.5%"  "4.4%"  "2.9%"  "0.1%"  "0.1%"  "1.6%"  "1.6%"

# Assign formats from the catalog to format attributes
formats(df) <- cat1

# Apply formats
fdata(df)
#          state    area   pct
# 1      Alabama  51,609  1.4%
# 2       Alaska 589,757 16.3%
# 3      Arizona 113,909  3.1%
# 4     Arkansas  53,104  1.5%
# 5   California 158,693  4.4%
# 6     Colorado 104,247  2.9%
# 7  Connecticut   5,009  0.1%
# 8     Delaware   2,057  0.1%
# 9      Florida  58,560  1.6%
# 10     Georgia  58,876  1.6%

```
In normal use, of course, the format catalog would likely be created 
in a separate script and saved to a file using the `write.fcat()` function.
The format catalog can then be read by any number of programs using 
the `read.fcat()` function, and the formats in the catalog can be 
applied as needed to your data.  

### Next Steps

For additional reinforcement of the topics presented above, please read the 
following articles:

* [Format Data Function](fmtr-fdata.html)
* [Format Apply Function](fmtr-fapply.html)
* [Format Catalogs](fmtr-fcat.html)
* [Format and Combine](fmtr-fapply2.html)
* [Convenience Functions](fmtr-convenience.html)
* [Helper Functions](fmtr-helpers.html)
* [Complete Example 1](fmtr-example1.html)
* [Complete Example 2](fmtr-example2.html)

