---
title: "Format Data Function"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Format Data Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### How to use fdata()
Data can be formatted by assigning formats to the **format** attribute
of the columns in your dataframe or tibble, and then by calling the `fdata()` 
function on that data.  A sample program is as follows:

```{r eval=FALSE, echo=TRUE}
# Set up data frame
df <- mtcars[1:10, c("mpg", "cyl")]
df

# Define and assign formats
attr(df$mpg, "format") <- value(condition(x >= 20, "High"),
                                condition(x < 20, "Low"))

attr(df$cyl, "format") <- function(x) format(x, nsmall = 1)

# Apply formatting
fdata(df)

```

Here is the mtcars subset before formatting:
```
#                    mpg cyl
# Mazda RX4         21.0   6
# Mazda RX4 Wag     21.0   6
# Datsun 710        22.8   4
# Hornet 4 Drive    21.4   6
# Hornet Sportabout 18.7   8
# Valiant           18.1   6
# Duster 360        14.3   8
# Merc 240D         24.4   4
# Merc 230          22.8   4
# Merc 280          19.2   6
```

And here is the mtcars subset after formatting:
```
#                    mpg cyl
# Mazda RX4         High 6.0
# Mazda RX4 Wag     High 6.0
# Datsun 710        High 4.0
# Hornet 4 Drive    High 6.0
# Hornet Sportabout Low  8.0
# Valiant           Low  6.0
# Duster 360        Low  8.0
# Merc 240D         High 4.0
# Merc 230          High 4.0
# Merc 280          Low  6.0

```

You may apply formatting to variables of any data type: character, numeric, 
date, etc. Internally, the `fdata()` function is using the `fapply()`
function on each column in the data frame.  If there is no format assigned
to a column, that column is returned unaltered.

